# Regression and model validation excercise

*Describe the work you have done this week and summarize your learning.*

- Describe your work and results clearly. 
- Assume the reader has an introductory course level understanding of writing and reading R code as well as statistical methods
- Assume the reader has no previous knowledge of your data or the more advanced methods you are using  


```{r, include = F}
# Libraries required:
library(pacman)
library(dplyr) 
library(GGally)
library(ggplot2)
```


```{r}
# Reading data:
learning2014 <- read.table("http://s3.amazonaws.com/assets.datacamp.com/production/course_2218/datasets/learning2014.txt", sep=",", header=TRUE)
```

The data here describes results from a questionaire done to students on a course "Johdatus yhteiskuntatilastotieteeseen", and combines the information with the background and exam points of each student. The questionaire aim to uncover effects of learning strategies, (_i.e._ deep, surface and strategic learning methods). Aspects covered in the questionaire were related to attitude towards statistics and the learning methods used as follows:

* Deep learning aspects (Seeking Meaning, Relating Ideas, and Use of Evidence) 
* Surface learning aspects (Lack of Purpose, Unrelated Memorising, and Syllabus-boundness)
* Strategic learning aspects (Organized Studying, and Time Management)

```{r}
# Structure of the data
str(learning2014)
```

```{r}
# Dimensions of the data
dim(learning2014)
```

Here is a graphical overview of the data. Majority od the students seem to be less than 25 years old. The female students represent 2 thirds of the students in question. The answers of female (red) and male (blue) students seem to differ mainly in the attitude and surface learning. From the answers, the attitude towards statistics seems to correlate best with the aexamination points. There is a negative correlation between deep and surface learning evident in the answers of male students (but not in the answers of female students).

```{r}
p <- ggpairs(learning2014, mapping = aes(col = gender, alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```

Let us try to model the data.

```{r}
# A linear model
my_model <- lm(points ~ attitude + age + gender, data = learning2014)
# Summary of the model
summary(my_model)
```

The residuals seem to be distributed somewhat equally around value 0 in the quantiles. The standard error for the intercept and attitude are aprox. one magnitude smaller than the estimate, their t-values are high and P-values are low (therefore the significance is good). Age and gender are not good explanatory variables to model the examination points. Therefore, let us remove those variables from the model. Now, the model says that if attitude increases by 1 unit, then the exam points increase by 3.5 points. The model can explain approx. 19 % of the variation in the modelled variable.

```{r}
# Remodelling without age and gender as explanatory variables:
my_model <- lm(points ~ attitude, data = learning2014)
# Summary of the model
summary(my_model)
```

Let us next turn to the graphical model validation. Our linear model assumes linear relationship between explanatory variable (attitude) and the modelled variable (examination points). The errors of the model are assumed to be normally distributed around zero and not correlated with each other. The normal Q-Q plot shows relatively good normal distribution of the residuals (only, the estimates with the lowest residuals are slightly too low). The residuals vs. fitted plot shows that exactness of the estimate does not depend on the amount of examination points. The residuals vs. leverage plot shows, that there are leverages of the individual points are low and that there are no evident trends visible. 

```{r}
# Plotting the graphical model validation images:
par(mfrow = c(2,2))
plot(my_model, which = c(1,2,5))
```


